# Dimensionality reduction techniques

```{r}
# access the packages
library(MASS); library(corrplot); library(tidyr); library(corrplot); library(dplyr); library(ggplot2); library(GGally); library(psych); library(DescTools); 

# read the data into memory 
human <- read.csv("C:/Users/juusov/Documents/IODS-project/Data/human.csv", row.names = 1)

# Explore the structure and the dimensions of the data
str(human)
colnames(human)
row.names(human)
describe(human)
summary(human)
```

## Description of 'human' dataset variables. 

The 'human' dataset originates from the United Nations Development Programme. Human Development Indicators and Indices povide an overview of key aspects of human development.The data combines several indicators from most countries in the world. This data (19 diffrent variables and 195 observations) includes following variables:

* **Country** = Country name
* **GNI = Gross** National Income per capita
* **Life.Exp** = Life expectancy at birth
* **Edu.Exp** = Expected years of schooling 
* **Mat.Mor** = Maternal mortality ratio
* **Ado.Birth** = Adolescent birth rate
* **Parli.F** = Percetange of female representatives in parliament
* **Edu2.F** = Proportion of females with at least secondary education
* **Edu2.M** = Proportion of males with at least secondary education
* **Labo.F** = Proportion of females in the labour force
* **Labo.M** = Proportion of males in the labour force
* **Edu2.FM** = Edu2.F / Edu2.M
* **Labo.FM** = Labo2.F / Labo2.M

```{r}
# Draw distributions and correlations
ggpairs(human, lower = list(continuous = "smooth_loess")) + theme_classic()
```

```{r}
# Draw correlation plot
cor(human)%>%corrplot(method="number", type='upper', diag = FALSE)
```

## Describption of the distributions of the variables and the relationships between them. 

Most of the variables are highly skewed. Only two of them are nearly normally distributed (“Edu.Exp” and “Parli.F”). The skewed distributions suggests that some transformations on variables could improve performance of variables in the models. There seeems to be many strong correlation coefficients and some weak correlation coefficients, especially Parli.F.

```{r}
# perform principal component analysis (with the SVD method)
pca_human_not_std <- prcomp(human)
sum_pca_human_not_std <- summary(pca_human_not_std)
pca_pr_not_std <- round(100*sum_pca_human_not_std$importance[2, ], digits = 3)
pca_pr_not_std
```


```{r}
# standardize the variables
human_std <- scale(human)

# perform principal component analysis (with the SVD method)
pca_human_std <- prcomp(human_std)
sum_pca_human_std <- summary(pca_human_std)
pca_human_std <- round(100*sum_pca_human_std$importance[2, ], digits = 3)
pca_human_std
```

```{r}
# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human_std)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
```

From above you can find principal component analysis (PCA) on the not standardized human (the first one) and with standardizing (the last one). As we 
 